Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-17 ind_files/01_Introduction.tex
Minor phrasing changes

Studies of neutral or near-neutral dynamics in population and quantitative genetics usually do take stochasticity seriously, explicitly modeling finite populations that follow stochastic dynamics. Unfortunately, the classic or standard stochastic models in both population genetics~\citep{fisher_genetical_1930,wright_evolution_1931, moran_random_1958, kimura_diffusion_1964} and quantitative genetics~\citep{crow_introduction_1970, lande_natural_1976} typically assume a fixed total population size and their validity is therefore rather restrictive since population sizes routinely fluctuate in the real world. Those models which study evolution in populations with non-constant size usually impose deterministic, usually phenomenological, rules that are independent of population composition for how the total population size must vary~\citep{kimura_probability_1974,ewens_probability_1967,otto_probability_1997, engen_fixation_2009, waxman_unified_2011}. Such models are somewhat artificial since demography and population size are forced to be independent quantities even though this is obviously not the case in natural populations, where population size is a `bulk' property whose value emerges from an intricate interplay of the individual-level demographic processes of birth and death. Notably, the Price equation is usually formulated in a deterministic, dynamically insufficient\footnote{For the non-theorists: `Dynamically insufficient' means that the equation cannot be iterated in the form of a difference equation $x_{t+1} = F(x_t)$ or a differential equation $\dot{x} = F(x)$  to predict a dynamic `trajectory' given an initial condition $x_0$. The term `insufficient' is to indicate that the equation requires `complete' information and only relates two quantities `in retrospect': In our case, in its most general setting, the Price equation is usually formulated in a way that partitions a \emph{given} amount of phenotypic change between two populations (usually the same population at two different times) into change due to selection, transmission bias, etc., rather than \emph{predicting} a trajectory for how much phenotypic change will occur at various future times based on the \emph{current} phenotypic distribution.} setting (but see~\cite{page_unifying_2002} and~\cite{lion_theoretical_2018} for a dynamically sufficient formulation that is deterministic and~\cite{rice_stochastic_2008} for a stochastic formulation that is dynamically insufficient). Since real-life populations are stochastic, finite, and of non-constant population size, this is somewhat of a problem, since we know that such deterministic approximations that may not capture important dynamics of the real system of interest - finite, fluctuating populations that vary stochastically.

@@

Chapter \ref{chap_math_background} provides the basic mathematical background required and illustrates the major ideas that we will use. To facilitate readership by a broad audience, I only assume passing familiarity with calculus (derivatives, integrals, Taylor expansions) and probability. Familiarity with stochastic calculus is helpful for some sections but is not required. I present a brief introduction to the relevant mathematics in section \ref{sec_math_background} and present a toy example of tracking population size of a population of identical individuals in section \ref{sec_1D_processes}. I introduce a description of the system via a `master equation', and then conduct a `system-size expansion' to obtain a Fokker-Planck equation for the system. Finally, I conduct a weak noise approximation to arrive at a linear Fokker-Planck equation which can be solved exactly using some stochastic calculus to arrive at a closed-form solution given by a time-dependent Ornstein-Uhlenbeck process, thus illustrating all the major tools required.

@@

While the mathematics of chapter \ref{chap_BD} is standard and well-understood, it has, to the best of my knowledge, not been used in this context in the generality we use here. Several specific models of specific systems do use these mathematical techniques, but these papers are often written assuming familiarity with notions in physics and/or mathematics and thus may not be very accessible to theoretical ecologists who do not have formal training in these subjects (but see~\cite{czuppon_understanding_2021} for a recent pedagogical review on the general approach applied to Wright-Fisher and Moran processes, where total population size is constant). As such, chapters \ref{chap_math_background} and \ref{chap_BD} together also serve as a tutorial and technical introduction to some theoretical ideas: For ecologists, the chapter introduces `system size expansions' and illustrates their use in a general setting, and can be seen as a tutorial on modelling finite populations analytically with minimal assumptions; For population geneticists, the chapter illustrates how system-size approximations (`diffusion approximations' in the population genetics literature) can be carried out without assuming a constant (effective) population size and how this generalization has important consequences for the evolutionary forces at play; For physicists and applied mathematicians, the chapter presents a study of the consequences of applying the system-size expansion to the kind of density-dependent birth-death processes that are widely applicable in ecology and evolution - Unlike many physical systems, though demographic processes like birth and death over ecological timescales are usually formulated in terms of population numbers or densities, predictions in evolution are typically in terms of frequencies of types, and this fact has subtle consequences that can be overlooked if one only works with densities.

>>>>

Studies of neutral or near-neutral dynamics in population and quantitative genetics usually do take stochasticity seriously, explicitly modeling finite populations that follow stochastic dynamics. Unfortunately, the classic or standard stochastic models in both population genetics~\citep{fisher_genetical_1930,wright_evolution_1931, moran_random_1958, kimura_diffusion_1964} and quantitative genetics~\citep{crow_introduction_1970, lande_natural_1976} typically assume a fixed total population size and their validity is therefore rather restrictive since population sizes routinely fluctuate in the real world. Those models which study evolution in populations with non-constant size usually impose deterministic, typically phenomenological, rules for how the total population size must vary~\citep{kimura_probability_1974,ewens_probability_1967,otto_probability_1997, engen_fixation_2009, waxman_unified_2011}. These rules also usually do not depend on population composition. Such models are thus somewhat artificial, since demography and population size are forced to be independent quantities even though this is obviously not the case in natural populations, where population size is a `bulk' property whose value emerges from an intricate interplay of the individual-level demographic processes of birth and death~\citep{metcalf_why_2007}. Notably, the Price equation is usually formulated in a deterministic, dynamically insufficient\footnote{Since this is not very standard nomenclature outside theoretical biology and related fields: `Dynamically insufficient' means that the equation cannot be iterated in the form of a difference equation $x_{t+1} = F(x_t)$ or a differential equation $\dot{x} = F(x)$  to predict a dynamic `trajectory' given an initial condition $x_0$. The term `insufficient' is to indicate that the equation requires `complete' information and only relates two quantities `in retrospect': In our case, in its most general setting, the Price equation is usually formulated in a way that partitions a \emph{given} amount of phenotypic change between two populations (usually the same population at two different times) into change due to selection, transmission bias, etc., rather than \emph{predicting} a trajectory for how much phenotypic change will occur at various future times based on the \emph{current} phenotypic distribution.} setting (but see~\cite{page_unifying_2002} and~\cite{lion_theoretical_2018} for a dynamically sufficient formulation that is deterministic and~\cite{rice_stochastic_2008} for a stochastic formulation that is dynamically insufficient). Since real-life populations are stochastic, finite, and of non-constant population size, this is somewhat of a problem, since we know that such deterministic approximations that may not capture important dynamics of the real system of interest - finite, fluctuating populations that vary stochastically.

@@

Chapter \ref{chap_math_background} provides the basic mathematical background required and illustrates the major ideas that we will use. To facilitate readership by a broad audience, I only assume passing familiarity with calculus (derivatives, integrals, Taylor expansions) and probability. Familiarity with stochastic calculus is helpful for some sections but is not required, and I present a brief introduction to the relevant notions from both Markov theory and stochastic calculus in section \ref{sec_math_background}. In section \ref{sec_1D_processes}, I present a toy example of tracking population size of a population of identical individuals in section. I introduce a description of the system via a `master equation', and then conduct a `system-size expansion' to obtain a Fokker-Planck equation for the system, thus illustrating all the major tools required. For completeness, I also conduct a weak noise approximation to arrive at a so-called `linear' Fokker-Planck equation that can be solved exactly to arrive at a closed-form solution.

@@

While the mathematics of chapter \ref{chap_BD} is standard and well-understood, it has, to the best of my knowledge, not been applied before in the generality and context we use here. Several specific models of specific systems do use these mathematical techniques, but these papers are often written assuming familiarity with notions in physics and/or mathematics and thus may not be very accessible to theoretical ecologists who do not have formal training in these subjects (but see~\cite{czuppon_understanding_2021} for a recent pedagogical review on the general approach applied to Wright-Fisher and Moran processes, where total population size is constant). As such, chapters \ref{chap_math_background} and \ref{chap_BD} together also serve as a tutorial and technical introduction to some theoretical ideas: For ecologists, the chapter introduces `system size expansions' and illustrates their use in a general setting, and can be seen as a tutorial on modelling finite populations analytically with minimal assumptions; For population geneticists, the chapter illustrates how system-size approximations (`diffusion approximations' in the population genetics literature) can be carried out without assuming a constant (effective) population size and how this generalization has important consequences for the evolutionary forces at play; For physicists and applied mathematicians, the chapter presents a study of the consequences of applying the system-size expansion to the kind of density-dependent birth-death processes that are widely applicable in ecology and evolution - Unlike many physical systems, though demographic processes like birth and death over ecological timescales are usually formulated in terms of population numbers or densities, predictions in evolution are typically in terms of frequencies of types, and this fact has subtle consequences that can be overlooked if one only works with densities, as is often done in physics studies of eco-evolutionary systems.

### 2023-03-17 ind_files/05_unification_discussion.tex
Minor phrasing changes

Equation \eqref{nD_eqn_for_frequencies}, which we derived in chapter \ref{chap_BD}, is a very general equation for how frequencies change over time in stochastic populations. To recap, we started with a population which can contain up to $m$ different types of individuals. The population as a whole is characterized by a vector $\mathbf{n}(t) = [n_1(t), \ldots, n_m(t)]$ indexing the number of individuals of each type in the population. Changes of the population are through either birth or death of single individuals. On ecological grounds, we postulated the existence of a `system-size' parameter $K>0$ that leads to density-dependent growth and prevents the population from growing infinitely large. We then moved from numbers $\mathbf{n}(t)$ to population densities $\mathbf{x}(t)$ by assuming that birth and death rates depend on population density $x$ and not only on population numbers $n$. In the regime where $K$ is not too small (corresponding to `medium sized' populations), we found a continuous description of how $x$ changes stochastically. We further assumed that each type has a per-capita birth rate $b^{\textrm{(ind)}}(\mathbf{x})$, a per-capita death rate $d^{\textrm{(ind)}}(\mathbf{x})$, and an additional term $\mu Q_{i}(\mathbf{x})$ representing mutational effects. All three of these functions may in general vary in an arbitrarily complicated frequency-dependent manner (or be frequency independent constants, of course). We then saw that $p_i$, the frequency of the $i\textsuperscript{th}$ type in the population, changes according to the stochastic differential equation:

>>>>

Equation \eqref{nD_eqn_for_frequencies}, which we derived in chapter \ref{chap_BD}, is a very general equation for how frequencies change over time in stochastic populations. To recap, we started with a population which can contain up to $m$ different (fixed) types of individuals. The population as a whole is characterized by a vector $\mathbf{n}(t) = [n_1(t), \ldots, n_m(t)]$ indexing the number of individuals of each type in the population. Changes of the population are through either birth or death of single individuals. On ecological grounds, we postulated the existence of a `system-size' parameter $K>0$ that leads to density-dependent growth and prevents the population from growing infinitely large. We then moved from numbers $\mathbf{n}(t)$ to population densities $\mathbf{x}(t)=\mathbf{n}/K$ by assuming that birth and death rates depend on population densities $\mathbf{x}$ and not only on population numbers $\mathbf{n}$. In the regime where $K$ is not too small (corresponding to `medium sized' populations), we found a continuous description of how $\mathbf{x}$ changes stochastically. We further assumed that each type has a per-capita birth rate $b^{\textrm{(ind)}}(\mathbf{x})$, a per-capita death rate $d^{\textrm{(ind)}}(\mathbf{x})$, and an additional term $\mu Q_{i}(\mathbf{x})$ representing mutational effects. All three of these functions may in general vary in an arbitrarily complicated frequency-dependent manner (or be frequency independent constants, of course). We then moved from type densities $\mathbf{x}(t)$ to type frequencies $\mathbf{p}(t)$ and saw that $p_i$, the frequency of the $i\textsuperscript{th}$ type in the population, obeys the stochastic differential equation:

### 2023-03-17 ind_files/06_nontechnical_discussion.tex
Minor phrasing changes

Indeed, the most important takeaway from this thesis is that the equations of section \ref{sec_fun_theorems} are very general, since part \ref{part_theory} makes essentially no assumptions other than density dependence, the impossibility of infinite growth starting from finite population size, and the ability to define per-capita birth and death rates, and thus, these birth and death rate function(al)s can in general be almost arbitrarily complicated. Further, as we saw in section \ref{sec_fun_theorems}, the terms of the equations lend themselves to simple biological interpretation. Like the classical Price equation, the utility of the equations of section \ref{sec_fun_theorems} lies not (necessarily) in their solutions for specific models, but instead in their generality and the fact that their terms help us clearly think about the various evolutionary phenomena operating in biological populations~\citep{frank_natural_2012,luque_one_2017, luque_mirror_2021}. The general spirit of this thesis is thus identical to the philosophy behind the general move towards trying to formulate `model-independent' eco-evolutionary theory, a move that has recently been (rapidly) gaining popularity in the literature~\citep{grafen_formal_2014, queller_fundamental_2017, lion_theoretical_2018, allen_mathematical_2019, rice_universal_2020, week_white_2021, wickman_theoretical_2022, kuosmanen_turnover_2022, mazzolini_universality_2022,lion_extending_2023,allen_natural_2023}. However, there are some biologically important factors that I have neglected in my formalism. My framework works with unstructured populations, neglecting any population structure due to groups, age, class, sex, developmental stage, or space\footnote{Of course,  since position in continuous space is just a special case of a quantitative trait, we have technically incorporated space in a very limited sense: The formalism of chapter \ref{chap_infD_processes} can equally well describe a finite population of exactly identical individuals moving through one-dimensional space, and so can technically describe phenomena like range expansion of clonal populations. However, I am only noting this as a technicality - Such populations are likely somewhat `evolutionarily boring', since all individuals in the model must always be exactly identical in all aspects other than spatial location.}, all of which can lead to very complex and often surprising dynamics. Explicitly incorporating features such as population structure, sex, and space from first principles in an analytically tractable framework is a formidable task that will likely need innovative new mathematical and biological arguments, and presents a fantastic opportunity for future studies.

Descriptions such as the classic Price equation and the equations I present in this thesis `abstract away' system-specific details and almost inevitably come at the cost of precision~\citep{levins_strategy_1966, potochnik_idealization_2018}. These approaches are thus intended to complement empirical studies and modelling approaches that carefully study specific systems. To quote Robert Millikan~\citep{millikan_electron_1924}, ``Science walks forward on two feet, namely theory and experiment [...] Sometimes it is one foot which is put forward first, sometimes the other, but continuous progress is only made by the use of both - by theorizing and then testing, or by finding new relations in the process of experimenting and then bringing the theoretical foot up and pushing it on beyond, and so on in unending alterations.''

>>>>

Indeed, the most important takeaway from this thesis is that the equations of section \ref{sec_fun_theorems} are very general, since part \ref{part_theory} makes essentially no assumptions other than density dependence, the impossibility of infinite growth starting from finite population size, and the ability to define per-capita birth and death rates, and thus, these birth and death rate function(al)s can in general be almost arbitrarily complicated. Further, as we saw in section \ref{sec_fun_theorems}, the terms of the equations lend themselves to simple biological interpretation. Like the classical Price equation, the utility of the equations of section \ref{sec_fun_theorems} lies not (necessarily) in their solutions for specific models, but instead in their generality and the fact that their terms help us clearly think about the various evolutionary phenomena operating in biological populations~\citep{frank_natural_2012,luque_one_2017, luque_mirror_2021}. The general spirit of this thesis is thus in line with the general idea of trying to formulate `model-independent' eco-evolutionary theory that has recently been (rapidly) gaining popularity in the literature~\citep{grafen_formal_2014, queller_fundamental_2017, lion_theoretical_2018, allen_mathematical_2019, rice_universal_2020, week_white_2021, wickman_theoretical_2022, kuosmanen_turnover_2022, mazzolini_universality_2022,lion_extending_2023,allen_natural_2023}. However, there are some biologically important factors that I have neglected in my formalism. My framework works with unstructured populations, neglecting any potential effects of groups, age, class, sex, developmental stage, or space\footnote{Of course,  since position in continuous space is just a special case of a quantitative trait, we have technically incorporated space in a very limited sense: The formalism of chapter \ref{chap_infD_processes} can equally well describe a finite population of exactly identical individuals moving through one-dimensional space, and so can technically describe phenomena like range expansion of clonal populations. However, I am only noting this as a technicality - Such populations are likely somewhat `evolutionarily boring', since all individuals in the model must always be exactly identical in all aspects other than spatial location.}, all of which can lead to very complex and often surprising dynamics. Explicitly incorporating features such as population structure, sex, and space from first principles in an analytically tractable framework is a formidable task that may need innovative new mathematical and biological arguments, and presents a fantastic opportunity for future studies.

Descriptions such as the classic Price equation and the equations I present in this thesis `abstract away' system-specific details and almost inevitably come at the cost of precision~\citep{levins_strategy_1966, potochnik_idealization_2018}. These approaches are thus intended to complement empirical studies and modelling approaches that carefully study specific systems and generate vital knowledge about how these systems behave. To quote Robert Millikan~\citep{millikan_electron_1924}, ``Science walks forward on two feet, namely theory and experiment [...] Sometimes it is one foot which is put forward first, sometimes the other, but continuous progress is only made by the use of both - by theorizing and then testing, or by finding new relations in the process of experimenting and then bringing the theoretical foot up and pushing it on beyond, and so on in unending alterations.''
